Fast radio burst cosmology: dispersion measure angular power spectrum

### 2015-01-26 writeup/dm_angular_ps.tex
Reorganized... again.

%\documentclass[onecolumn,amsmath,amssymb,floatfix,12pt,nofootinbib]{revtex4}
\documentclass[onecolumn,prd,noshowpacs,nofootinbib,amsmath,amssymb]{revtex4}
\usepackage{graphicx,epsfig,psfrag,bm,amssymb}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{mciteplus}
\usepackage{color}
\usepackage{mathrsfs,amsfonts,hepunits, color}
\usepackage{mciteplus}
\usepackage{tikz}
\usepackage{slashed}
\usepackage{verbatim}
\usetikzlibrary{arrows,shapes}
\usetikzlibrary{trees}
\usetikzlibrary{matrix,arrows} 				% For commutative diagram
											% http://www.felixl.de/commu.pdf
\usetikzlibrary{positioning}				% For "above of=" commands
\usetikzlibrary{calc,through}				% For coordinates
\usetikzlibrary{decorations.pathreplacing}  % For curly braces
\usepackage{pgffor}							% For repeating patterns

\usetikzlibrary{decorations.pathmorphing}	% For Feynman Diagrams
\usetikzlibrary{decorations.markings}
\usetikzlibrary{snakes}

\tikzset{
	% >=stealth', %%  Uncomment for more conventional arrows
    vector/.style={decorate, decoration={snake}, draw},
	provector/.style={decorate, decoration={snake,amplitude=2.5pt}, draw},
	antivector/.style={decorate, decoration={snake,amplitude=-2.5pt}, draw},
    fermion/.style={draw, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw]{>}}}},
    fermionbar/.style={draw, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw=black]{<}}}},
    fermionnoarrow/.style={draw},
    gluon/.style={decorate, draw,decoration={coil,amplitude=4pt, segment length=6pt}, line width=1},
    scalar/.style={dashed,draw, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw]{>}}}},
    scalarbar/.style={dashed,draw, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw]{<}}}},
    scalarnoarrow/.style={dash pattern = on 6 pt off 3 pt,draw},
    electron/.style={draw, postaction={decorate},
        decoration={markings,mark=at position .55 with {\arrow[draw]{>}}}},
	bigvector/.style={decorate, decoration={snake,amplitude=4pt}, draw},
	vectorscalar/.style={loosely dotted,draw, postaction={decorate}},
}
\RequirePackage{xspace}
\usepackage{relsize}
\usepackage{slashed}
\newcommand{\fb}{{\rm fb}}
\newcommand{\ab}{{\rm ab}}

\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\lsim{\mathrel{\rlap{\lower4pt\hbox{\hskip 0.5 pt$\sim$}}
    \raise1pt\hbox{$<$}}}                % less than or approx. symbol
\def\gsim{\mathrel{\rlap{\lower4pt\hbox{\hskip1pt$\sim$}}
    \raise1pt\hbox{$>$}}}

\newcommand{\schi}{s_{\chi\bar\chi}}  % tried to use this throughout off-shell derivation so that we can change notation later if desired.
\newcommand{\scratch}[1]{{\textcolor{red}{ #1}}}
\newcommand{\f}{\frac}
\newcommand{\pf}[2]{\left(\frac{#1}{#2}\right)}
\newcommand{\g}{{\rm g}}
\newcommand{\s}{{\rm s}}
\newcommand{\m}{{\rm m}}
\def\lsim{\mathrel{\rlap{\lower4pt\hbox{\hskip1pt$\sim$}}
    \raise1pt\hbox{$<$}}}
\def\gsim{\mathrel{\rlap{\lower4pt\hbox{\hskip1pt$\sim$}}
    \raise1pt\hbox{$>$}}}
\newcommand{\vev}[1]{ \left\langle {#1} \right\rangle }
\newcommand{\bra}[1]{ \langle {#1} | }
\newcommand{\ket}[1]{ | {#1} \rangle }
\newcommand{\ev}{ {\rm eV} }
\newcommand{\kev}{{\rm keV}}
\newcommand{\mev}{{\rm MeV}}
\newcommand{\gev}{{\rm GeV}}
\newcommand{\tev}{{\rm TeV}}
\newcommand{\mpl}{$M_{Pl}$}
\newcommand{\mw}{$M_{W}$}
\newcommand{\Ft}{F_{T}}
\newcommand{\Zparity}{\mathbb{Z}_2}
\newcommand{\BLambda}{\boldsymbol{\lambda}}

% shortcuts
\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\bD}{\boldsymbol D}
\newcommand{\bC}{\boldsymbol C}
\newcommand{\bDelta}{\boldsymbol \Delta}
\newcommand{\Dgal}{D^{\rm gal}}
\newcommand{\Dne}{D^{\bar{n}_e}}
\newcommand{\Dde}{D^{\delta_e}}
\newcommand{\Dsrc}{D^{\rm src}}
\newcommand{\del}{\delta\!}
\newcommand{\calP}{{\cal P}}

@@

\author{Kiyoshi Wesley Masui}
 \affiliation{Department of Physics and Astronomy, University of British 
Columbia, Vancouver, BC, Canada V6T 1Z1}

\begin{abstract}
In recent years a new astrophysical phenomena, fast radio bursts (FRBs), has been detected that potentially originates from cosmological distances.  These broadband radio bursts  typically last for only milliseconds and are a \emph{standard ping} that can be used to map the properties of the Universe.  In this paper,  under the assumption that these bursts originate from cosmological distances, we discuss how the detection of a large enough population of these events can be used to map the cosmological properties of the Universe.    We first discuss one point statistics and show that, even if FRBs are not standard candles, the dispersion-measure distance vs. luminosity-distance distribution of any observed population is a strong constraint on cosmology.  We then discuss how the two point statistics of an observed population of FRBs traces the large-scale structure of the Universe and make predictions for the angular power spectra of dispersion-measure distance.  It is possible that FRBs do not originate at cosmological distances, but if they do, we argue that they could open a remarkable window to observe the properties of the Universe.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%					Introduction
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Several years ago an unusual radio event was discovered at the Parks radio telescope while searching for millisecond pulsars.  In a direction out of the plane of the Galaxy an event was discovered with an unusually large dispersion-measure, implying a column of electrons much large than expected in the Galaxy.

@@

The outline of this \emph{Letter} is as follows. In Section \ref{sec:notes} we discuss how dispersion measure provides a new distance measure in cosmology and how the joint distribution of dispersion-distance and luminosity-distance of a population of FRBs depends on cosmology and the intrinsic luminosity distance distribution.   In Section \ref{sec:notes} we then show how to calculate the angular power spectra of dispersion distance.  We then forecast how well these cosmological observables can be measured.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%					Basic Ingredients 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%Let's define the dispersion-measure distance.

%The lets discuss how a given luminosity relationship and cosmology dictates an expected joint distribution of dispersion measure.

@@

We define the dispersion-measure distance (or dispersion distance),
the equivalent distance required to
produce a given amount of dispersion in an observed radio signal in the
Universe today, as $ D = {{\rm DM}}/{\bar{n}_{e0}}$
We separate the dispersion distance into the contributions from each
term in \ref{e:ne_sep}:
\begin{equation}
\label{e:D_sep}
D = \Dgal + \Dne + \Dde + \Dsrc
\end{equation}
For a population of FRBs originating at cosmological distances, ${\rm DM}^{\rm
src}$ is only a single function of $\hat{n}$ which can presumably be well
measured.  In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, we will
here-after we ignore this term.

@@

We are principally interested in the $D^{\delta_e}$ term as it is a probe of
large-scale structure. In the next section we relate the statistics of this
term to cosmology.

@@

\section{Dispersion measure angular correlations}

@@

@@

\begin{equation}
C_l^{zz'}=\frac{2}{\pi} \int_0^{\infty}k^2 dk P_{ee}(k) \int_0^{\eta^z}\frac{d\eta}{a(\eta)} \int_0^{\eta^{z'}}\frac{d\eta'}{a(\eta')}  j_l(k\eta) j_l(k\eta')
\end{equation}
\begin{equation}
C_l^{\Dde \Dde} = \int d\eta^z \int d\eta^{z'} {\cal P}(\eta^z){\cal P}(\eta^{z'}) C_l^{z z'}
\end{equation}

@@

The joint probablility of observing a pair of FRB events at 3D coordinates
$\vec x_i$ and $\vec x_j$ in infintessimal volumes $\ud^3\vec x_i$ and
$\ud^3\vec x_j$ is porportional to
\begin{equation}
    \calP(\vec x_i, \vec x_j) = \calP(\eta_i) \calP(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \frac{\ud^3\vec x_i}{\eta_i^2}
    \frac{\ud^3\vec x_j}{\eta_j^2}.
\end{equation}
The factors of $1/\eta^2$ are nessisary to achive the proper volume weighting,
a fact that can be seen by taking the limit of $\xi \to 0$ and integrating of
the angular coordinates. Noting that $\calP(\eta_i)$ has units 
$({\rm MPc}\,{\rm rad}^2)^{-1}$.

An observed event has known angular coordinates, and we are interested in
correlations in $\Dne$ which is directly related to the line of sight
coordinate.  Fixing the angular coordinates:
\begin{equation}
    \calP(\eta_i, \eta_j | \hat{n}_i, \hat{n}_i) = N(\hat{n}_i, \hat{n}_i)
    \calP(\eta_i)\calP(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \ud \eta_i \ud \eta_j,
\end{equation}
where $N(\hat{n}_i, \hat{n}_i)$ is a normalizing factor which ensures the
expression integrates to unity over $\eta_i$ and $\eta_j$. The factors of
$1/\eta^2$ have been canceled when converting $\ud^3 \vec x = \eta^2 \ud \eta \ud^2
\vec \theta$. If the second term in brackets is small, then the normalizations
$N$ is 1 (since the distributions $\calP{\eta_i}$ are already normalized. 
This expression is
only valid if $i \ne j$.
Transforming to $\Dne$ as our distance measure, we obtain
\begin{align}
\calP(\Dne_i, \Dne_j | \hat{n}_i, \hat{n}_i)
    &= \calP(\eta_i)\calP(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \eta_i}{\ud \Dne_i} \frac{\ud \eta_j}{\ud \Dne_j}
        \ud \Dne_i\ud \Dne_j
        \\
    &= \calP(\eta_i)\calP(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \Dne_i}{1 + z_i}\frac{\ud \Dne_j}{1 + z_j}
\end{align}
Finally, taking the second moment and subtracting off the product of the first
moments:
\begin{equation}
\langle \Dne_i \Dne_j \rangle
    - \langle \Dne_i \rangle \langle \Dne_j \rangle =
    \int \Dne_i \Dne_j
    \frac{\calP(\eta_i)}{1 + z_i}\frac{\calP(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\vec x_i - \vec x_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}

\section{Notes}\label{sec:notes}

\subsection{2D statistics}
% Need beter section title.

Here we consider two dimentional statistics where the cosmological information
lies in the correlations between dispersion measures at a given angular
separation.  The information that is in principle available in the angular
clustering of the FRB events is ignored.

As a cosmological probe this almost perfectly analogouse to the weak lensing 
of galaxies.
We wish to estimate the line of site electron column density,
which, like lensing shear, is an integral along the line-of-site to a source at an
uncertain redshift. Our observerable the dispersion measure, like galaxy
elipticity, gives a noisy estimate.  In both cases the angular clustering of
the sources is ignored.

The dispersion measures of two different FRBs are correlated by two effects.
The first is correlations in the intervening line-of-site large-scale
structure.  This correlates the $D^{\delta_e}$ part of the dispersion measures.
The second effect is that two FRBs with small angular separation are likely to
be at small redshift separations if FRB events trace large-scale structure.
This correlates the $D^{\bar{n}_e}$ of the two FRBs.  In principle all
permutations of $D^{\bar{n}_e}$, $D^{\delta_e}$ and $D^{\rm src}$ should be
correlated to some degree, however we believe that the two effect described
here are the dominatnt contributions and defer investications of other effect
to future work.

\subsubsection{The dispersion measure angular power-spectrum}

\subsubsection{estimating the dispersion measure angular power-spectrum}

\subsubsection{sensitivity estimates}

\subsection{Random notes about correlation function normalization}

\begin{equation}
\langle D_i \rangle =  \langle D^{\rm src}_i \rangle
    + \langle D^{\bar{n}_e}_i \rangle
\end{equation}

If $i \neq j$:
\begin{align}
\langle D_i D_j \rangle - \langle D_i \rangle \langle D_j \rangle
    =& \langle (D^{\delta_e}_i + D^{\rm src}_i +  D^{\bar{n}_e}_i)
              (D^{\delta_e}_j + D^{\rm src}_j +  D^{\bar{n}_e}_j) \rangle
              - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle [
        D^{\delta_e}_i
        + (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
        + \langle D^{\rm src}_i \rangle
        +  (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
        + \langle D^{\bar{n}_e}_i \rangle
    ]
    \nonumber \\
    & \qquad [
        D^{\delta_e}_j
       + (D^{\rm src}_j - \langle D^{\rm src}_j \rangle)
        + \langle D^{\rm src}_j \rangle
        +  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle)
        + \langle D^{\bar{n}_e}_j \rangle
    ] \rangle
    \nonumber \\
    & \qquad - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle D^{\delta_e}_i D^{\delta_e}_j \rangle
        + \langle (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
                 (D^{\rm src}_j - \langle D^{\rm src}_j \rangle) \rangle
        + \langle (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
                  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle) \rangle
    \nonumber \\
    & \qquad
        + \langle D^{\rm src}_i \rangle \langle D^{\rm src}_j \rangle
        + \langle D^{\bar{n}_e}_i \rangle \langle D^{\bar{n}_e}_j \rangle
        + \langle D^{\rm src}_i \rangle \langle D^{\bar{n}_e}_j \rangle
        + \langle D^{\bar{n}_e}_i \rangle \langle D^{\rm src}_j \rangle
    \nonumber \\
    & \qquad - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle D^{\delta_e}_i D^{\delta_e}_j \rangle
        + \langle (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
                 (D^{\rm src}_j - \langle D^{\rm src}_j \rangle) \rangle
        + \langle (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
                  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle) \rangle
\end{align}
In the last equation, the first term is the term we've been dealing with, the
second term is hopefully zero, and the third term is what Ue-Li is worried
about, which we still have to calculate. I don't see anything unexpeced.

\subsection{Relating dispersion measure angular power-spectrum to cosmology}

The angular correlation function of dispersion distance between two lines of sight at redshift $z$ and direction $\hat{n}$ and redshift $z'$ and direction $\hat{n}'$ is
\begin{equation}
w^{z z'}(\hat{n}\cdot\hat{n}') = \langle D^{\delta_e}(\hat{n}, z) D^{\delta_e}(\hat{n}', z') \rangle
\end{equation}
and the corresponding angular power spectrum is
\begin{equation}
w^{z z'}(\hat{n}\cdot\hat{n}') = \sum_l \frac{2l+1}{4 \pi} C_l^{z z'} P_l(\hat{n}\cdot\hat{n}')
\end{equation}

We can decompose the dispersion-distance field into spherical harmonics as
\begin{equation}
    D^{\delta_e}(\hat{n}, z) = \sum_{l m} D^{\delta_e}_{l m}(z) Y_{l m} (\hat{n})
\end{equation}

A plane wave can be expanded in terms of spherical eigenfunctions using the Rayleigh expansion
\begin{equation}
e^{i \vec{k}\cdot \hat{n} \eta} = 4\pi \sum_{l m} i^l j_{l}(k\eta) Y^*_{lm}(\hat{k})Y_{lm}(\hat{n})
\end{equation}
If we define 
\begin{equation}
n_e(\vec{k}) = \int d^3 \vec{x} e^{-i \vec{x}\cdot\vec{k}} n_e(\vec{x})
\end{equation}
to be the Fourier transform of the comoving electron density field then we can write
\begin{equation}
    D^{\delta_e}_{l m}(z) = \frac{4 \pi i^{l}}{\bar{n}_{e0}} \int \frac{d^3
    \vec{k}}{(2\pi)^3} n_e(\vec{k}) Y^*_{l m}(\hat{k}) \int_0^{\eta^z} d\eta a^{-1}(\eta) j_{l}(k\eta)
\end{equation}
and the angular power spectra of dispersion-distance, assumed to be an isotropic random field, is thus
\begin{equation}
    \langle D^{\delta_e}_{l m}(z) D^{\delta_e}_{l' m'}(z') \rangle = C_l^{z z'} \delta_{l l'} \delta_{m m'}
\end{equation}
Now since we have 
\begin{equation}
\langle n_e(\vec{k}) n_e(\vec{k}')\rangle = \bar{n}_{e0}^2\left[\langle \delta_e(\vec{k})\delta_e(\vec{k}')\rangle + 1 \right]
\end{equation}
where
\begin{equation}
\langle \delta_e(\vec{k})\delta_e(\vec{k}')\rangle = (2\pi)^3 \delta_{D}^3(\vec{k}-\vec{k}') P_{ee}(k)
\end{equation}
we finally have
\begin{equation}
C_l^{zz'}=\frac{2}{\pi} \int_0^{\infty}k^2 dk P_{ee}(k) \int_0^{\eta^z}\frac{d\eta}{a(\eta)} \int_0^{\eta^{z'}}\frac{d\eta'}{a(\eta')}  j_l(k\eta) j_l(k\eta')
\end{equation}
The one-point variance in the dispersion-distance from a redshift $z$ is just
\begin{equation}
(\sigma^2)^{zz}=\sum_{l} \frac{2l+1}{4\pi} C_l^{zz}
\end{equation}
Using the addition identify for a complete sum over spherical bessel functions we find that
\begin{equation}
(\sigma^2)^{zz}= \frac{1}{2\pi^2} \int_0^{\infty}k^2 dk P_{ee}(k) \int_0^{\eta^z}\frac{d\eta}{a(\eta)} \int_0^{\eta^{z}}\frac{d\eta'}{a(\eta')}  j_0(k\eta-k\eta')
\end{equation}

Given a source-population redshift distribution ${\cal P}(\eta^z)$, where 
\begin{equation}
\int d\eta^z {\cal P}(\eta^z) = 1
\end{equation}

The observable power spectrum of mean dispersion distance $\bar{D}$ is then
\begin{equation}
C_l^{\bar{D}\bar{D}} = \int d\eta^z \int d\eta^{z'} {\cal P}(\eta^z){\cal P}(\eta^{z'}) C_l^{z z'}
\end{equation}

We use the halo model to calculate $P_{ee}(k)$ and a potential ${\cal P}(z)$.

\subsection{Estimating the DM angular power spectrum from a sample of observed
FRBs}

To relate observations to cosmology, we wish to determine how a set of observed
dispersion measures $\boldsymbol D$ (whose members are $D_i$)
with known angular positions, $\hat n_i$, but with unknown
redshifts, $z_i$, depends on cosmological parameters.  We begin by defining the
redshift probability distribution function $p(\boldsymbol z)$. We assume that the
redshifts of the FRB sources are independent, \emph{i.e.}~we assume
\begin{equation}
    p(\boldsymbol z) = \prod_i p(z_i).
\end{equation}
This may not be strictly true since FRBs are expected to cluster and trace the
large-scale structure. This is a potential source of bias in the estimator we
derive.  From this assumption it follows that the $D^{\bar{n}_e}_i$ have
uncorrelated probability distributions
\begin{align}
    p(D^{\bar{n}_e}_i)
    &= p(z_i) \left|\frac{\ud D^{\bar{n}_e}}{\ud z} \right|_{z_i}^{-1} \\
        &= p(z_i) \left(\frac{\ud \eta}{\ud z} \right)_{z_i}^{-1},
\end{align}
whose second moment is
\begin{align}
    \label{e:sig_nebar}
    \langle D^{\bar{n}_e}_i D^{\bar{n}_e}_j \rangle
        &- \langle D^{\bar{n}_e}_i \rangle \langle D^{\bar{n}_e}_j \rangle \\
        &= \delta_{ij} \left[ \int_0^\infty \ud D^{\bar{n}_e} (D^{\bar{n}_e})^2
            p(D^{\bar{n}_e})
            - \left( \int_0^\infty \ud D^{\bar{n}_e} D^{\bar{n}_e}
        p(D^{\bar{n}_e}) \right)^2 \right]\\
        &\equiv \delta_{ij} \sigma_{\bar{n}_e}^2
\end{align}

We further assume that $D^{\rm src}_i$ are uncorrelated. Again this may not
be strictly true due to clustering of FRB sources and again this may bias our
estimator. We parameterized the variance from this term as 
\begin{equation}
    \label{e:sig_src}
    \langle D^{\rm src}_i D^{\rm src}_j \rangle
    - \langle D^{\rm src}_i \rangle \langle D^{\rm src}_j \rangle
    = \delta_{ij} \sigma^2_{\rm src}.
\end{equation}

\red{$D^{\bar{n}_e}_i$ and $D^{\rm src}_i$ are likely correlated due to
    redshift evolution of source events, but this does not affect the
    measurement since both are treated as noise terms.}

>>>>

\documentclass[onecolumn,prd,noshowpacs,nofootinbib,amsmath,amssymb]{revtex4}
\usepackage{graphicx,epsfig,psfrag,bm,amssymb}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{mciteplus}
\usepackage{color}
\usepackage{mathrsfs,amsfonts,hepunits, color}
\usepackage{mciteplus}
\usepackage{tikz}
\usepackage{slashed}
\usepackage{verbatim}

% shortcuts
%\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\bD}{\boldsymbol D}
\newcommand{\bC}{\boldsymbol C}
\newcommand{\bDelta}{\boldsymbol \Delta}
\newcommand{\Dgal}{D^{\rm gal}}
\newcommand{\Dne}{D^{\bar{n}_e}}
\newcommand{\Dde}{D^{\delta_e}}
\newcommand{\Dsrc}{D^{\rm src}}
\newcommand{\del}{\delta\!}
\newcommand{\calP}{{\cal P}}
\newcommand{\ud}{\,\mathrm{d}}

@@

\author{Kiyoshi Wesley Masui}
 \affiliation{Department of Physics and Astronomy, University of British 
Columbia, Vancouver, BC, Canada V6T 1Z1}
 \affiliation{Canadian Institute for Advanced Research, CIFAR Program in
Cosmology and Gravity, Toronto, ON, M5G 1Z8}

\begin{abstract}

\end{abstract}

@@

\section{Introduction}

Several years ago an unusual radio event was discovered at the Parks radio telescope while searching for millisecond pulsars.  In a direction out of the plane of the Galaxy an event was discovered with an unusually large dispersion-measure, implying a column of electrons much large than expected in the Galaxy.

@@

@@

@@

We define the dispersion-measure distance (or dispersion distance),
the equivalent distance required to
produce a given amount of dispersion in an observed radio signal in the
Universe today, as $ D = {{\rm DM}}/{\bar{n}_{e0}}$
We separate the dispersion distance into the contributions from each
term in Equation~\ref{e:ne_sep}:
\begin{equation}
\label{e:D_sep}
D = \Dgal + \Dne + \Dde + \Dsrc
\end{equation}
For a population of FRBs originating at cosmological distances, $\Dgal$
is only a single function of $\hat{n}$ which can presumably be well
measured.  In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, we will
here-after we ignore this term.

@@

@@

\section{Dispersion Measure angular correlations}

@@

We assume that $D^{\rm src}_i$ are uncorrelated. Again this may not
be strictly true due to clustering of FRB sources and environmental effects
and this may bias results.  In addition $\Dsrc$ and $\Dne$ may be correlated due
to redshift evolution of source events.

@@

\begin{equation}
C_l^{zz'}=\frac{2}{\pi} \int_0^{\infty}k^2 dk P_{ee}(k) \int_0^{\eta^z}\frac{d\eta}{a(\eta)} \int_0^{\eta^{z'}}\frac{d\eta'}{a(\eta')}  j_l(k\eta) j_l(k\eta')
\end{equation}
\begin{equation}
C_l^{\Dde \Dde} = \int d\eta^z \int d\eta^{z'} p(\eta^z)p(\eta^{z'}) C_l^{z z'}
\end{equation}

@@

The joint probablility of observing a pair of FRB events at 3D coordinates
$\vec x_i$ and $\vec x_j$ in infintessimal volumes $\ud^3\vec x_i$ and
$\ud^3\vec x_j$ is
\begin{equation}
    \calP(\vec x_i, \vec x_j) = p(\eta_i) p(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \frac{\ud^3\vec x_i}{\eta_i^2}
    \frac{\ud^3\vec x_j}{\eta_j^2}.
\end{equation}
The factors of $1/\eta^2$ are nessisary to achive the proper volume weighting,
a fact that can be seen by taking the limit of $\xi \to 0$ and integrating of
the angular coordinates. Noting that $p(\eta_i)$ has units 
\red{$({\rm MPc}\,{\rm rad}^2)^{-1}$ (conflicts with being normalized)}.

An observed event has known angular coordinates, and we are interested in
correlations in $\Dne$ which is directly related to the line of sight
coordinate.  Fixing the angular coordinates:
\begin{equation}
    \calP(\eta_i, \eta_j | \hat{n}_i, \hat{n}_i) = N(\hat{n}_i, \hat{n}_i)
    p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \ud \eta_i \ud \eta_j,
\end{equation}
where $N(\hat{n}_i, \hat{n}_i)$ is a normalizing factor which ensures the
expression integrates to unity over $\eta_i$ and $\eta_j$. The factors of
$1/\eta^2$ have been canceled when converting $\ud^3 \vec x = \eta^2 \ud \eta \ud^2
\vec \theta$. If the second term in brackets is small, then the normalizations
$N$ is 1 (since the distributions $p{\eta_i}$ are already normalized. 
This expression is
only valid if $i \ne j$.
Transforming to $\Dne$ as our distance measure, we obtain
\begin{align}
\calP(\Dne_i, \Dne_j | \hat{n}_i, \hat{n}_i)
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \eta_i}{\ud \Dne_i} \frac{\ud \eta_j}{\ud \Dne_j}
        \ud \Dne_i\ud \Dne_j
        \\
    &= p(\eta_i)p(\eta_j) \left[ 1 + b(z_i)
        b(z_j) \xi(\vec x_i - \vec x_j) \right]
        \frac{\ud \Dne_i}{1 + z_i}\frac{\ud \Dne_j}{1 + z_j}
\end{align}
Finally, taking the second moment and subtracting off the product of the first
moments:
\begin{equation}
\langle \Dne_i \Dne_j \rangle
    - \langle \Dne_i \rangle \langle \Dne_j \rangle =
    \int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\vec x_i - \vec x_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
And thus we have:
\begin{equation}
    w^{\bar{n}_e \bar{n}_e}(\hat{n}_i\cdot\hat{n}_j)
        =\int \Dne_i \Dne_j
    \frac{p(\eta_i)}{1 + z_i}\frac{p(\eta_j)}{1 + z_j}
    b(z_i) b(z_j) \xi(\eta_i \hat{n}_i - \eta_j \hat{n}_j)
        \ud \Dne_i\ud \Dne_j
\end{equation}
Where $\eta$ and $z$ are understood to be functions of $\Dne$.

\section{Modeling angular correlations}

\section{Estimating angular correlations}

We observe a set of observed dispersion measures
$\boldsymbol \Delta$ (whose members are $\delta D_i$)
with known angular positions, $\hat n_i$, but with unknown
redshifts, $z_i$.  The Covariance matrix is
${\boldsymbol C} \equiv
\langle \boldsymbol \Delta \boldsymbol \Delta^T \rangle$.
The off diagonal elements depend on cosmology
$C_{ij} = w(\hat{n}_i\cdot\hat{n}_j)$ while the diagonal part is dominated by noise.

\subsection{Noise}

The noise will be dominated by contributions from $\Dsrc$, $\Dne$ and $\Dde$.

\begin{align}
    \label{e:sig_nebar}
    \langle \delta D^{\bar{n}_e}_i \delta D^{\bar{n}_e}_i \rangle
        &= \left[ \int_0^\infty \ud D^{\bar{n}_e} (D^{\bar{n}_e})^2
            p(D^{\bar{n}_e})
            - \left( \int_0^\infty \ud D^{\bar{n}_e} D^{\bar{n}_e}
        p(D^{\bar{n}_e}) \right)^2 \right]\\
        &\equiv  \sigma_{\bar{n}_e}^2
\end{align}

\begin{equation}
    \label{e:sig_src}
    \langle \delta D^{\rm src}_i \delta D^{\rm src}_i \rangle
    \equiv \sigma^2_{\rm src}.
\end{equation}

\begin{equation}
    \label{e:sig_delta}
    \langle \delta \Dde_i \delta \Dde_i \rangle
		= \int d\eta^z \int d\eta^{z'} \delta(\eta^z - \eta^{z'}) 
		p(\eta^{z}) w^{z z'}(0)
\end{equation}

\subsection{Estimators}

Here we sketch how the dispersion mearure angular power-spectrum, $C_l$
could be estimated from a sample of observed dispersion-measures.

The covariance matrix of the sample is:

\begin{align}
    C_{ij} \equiv& \langle \delta D_i \delta D_j \rangle \\
    =&
        \begin{cases}
            \sum_l \frac{2l+1}{4 \pi} C_l P_l(\hat{n}_i\cdot\hat{n}_j)
                &  \text{if } (i \neq j) \\
            \sigma_{\bar{n}_e}^2 + \sigma^2_{\rm src}
                + \textrm{(the $w$ term)}
                &  \text{if } (i = j)
        \end{cases}
\end{align}

We see that the off diagonal terms depend on the angular power-spectrum of
interest while the diagonal is contaminiated by the unknown parameter
$\sigma_{\rm src}^2$. This is again in analogue to weak lensing where the
correlations between galaxy elipticty are caused primarily by cosmic leansing shear
while an individual galaxies elipticity is dominated the galaxy's intrinsic
elipticity. To remove the information in the diagonal, we define a
modified covariance matrix derivative such that:
\begin{equation}
    \tilde{C}_{ij,\alpha} \equiv
\begin{cases}
    C_{ij,\alpha} & \text{if } i \neq j,\\
0 & \text{if } i = j
\end{cases}.
\end{equation}

We will parameterize the angular power spectrum using band powers, such that:
\begin{equation}
    C_l \approx c_\alpha \quad \textrm{for $l^{min}_\alpha \leq l <
        l^{min}_{\alpha +1}$}
\end{equation}
giving
\begin{equation}
    C_{ij, \alpha} = \sum_{l = l^{min}_\alpha}^{l^{min}_{\alpha +1} - 1}
        \frac{2l+1}{4 \pi} P_l(\hat{n}_i\cdot\hat{n}_j)
        \quad\textrm{for $i \neq j$}.
\end{equation}

The estimator is then

\begin{equation}
    \hat{c}_\alpha = \frac{\tilde F^{-1}_{\alpha\beta}}{2} 
        \left[ \bDelta^T \bC^{-1} \tilde\bC_{,\beta} \bC^{-1} \bDelta \right],
\end{equation}
with
\begin{equation}
    \tilde F_{\alpha \beta} = \frac{1}{2} \mathrm{Tr} 
        \left[ \tilde \bC_{,\alpha} \bC^{-1} \tilde\bC_{,\beta} \bC^{-1}
        \right].
\end{equation}

\subsection{Sensitivity Estimates}

\section{Extensions to 3D}

\subsection{Dispersion measure space distortions}

\subsection{Four point statistics}

\section{Conclusions}
